---
title: "Pension Wealth Modeling"
author: "Anil"
date: "1/22/2021"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#equations: https://bookdown.org/yihui/bookdown/bookdown.pdf
```

This R Markdown document explains how Defined Benefits accrue by both text & interactive charts.
So, each key concept is explained, followed by embedded Shiny app that allows you to change inputs & produce new estimates.

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Final Average Salary (FAS)
https://anil-niraula.shinyapps.io/PensionWealthModeling/

```{r, echo=FALSE, error=FALSE, warning=FALSE, message = FALSE}
rm(list=ls())
###Load/install packages
#R.Version()
#https://github.com/ReasonFoundation/pensionviewr
#Create token -> usethis::edit_r_environ() -> restart -> Sys.getenv("GITHUB_PAT")
#install.packages('devtools')
#library(devtools)
#devtools::install_github("ReasonFoundation/reasontheme",force = TRUE)
#devtools::install_github("ReasonFoundation/pensionviewr", force = TRUE)
library(reasontheme, war)
library(pensionviewr)
library(ggplot2)
library(tidyverse)
library(tseries)
library(data.table)
library(readr)
#library(janitor)
library(tidyverse)
library(plyr)
#library(openxlsx)
library(tseries)
#library(ggplot2)
library(data.table)
#library(openxlsx)
#library(readr)
library(rsconnect)
library(base64enc)
#Shiny-----------
library(shiny)
library(shinyWidgets)
library(shinymanager)
library(repmis)
#library(shinyFiles)
library(DT)
library(plotly)


data <- pullStateData(2001)
    data <- filterData(data, 2017)
    data <- data %>% select(year, plan_name, state, arr,wage_inflation,payroll_growth_assumption,ee_nc_pct)
    
    
    
  #### Final Average Salary function ####
    fas <- function(arr, payroll_growth_assumption, start.wage, 
                max.yos, hire.age, smooth){
      
    ## create table w/ key inputs
     manual <- data.table("vesting" = 5, 
                       "smooth" = smooth, 
                       "hire.age" = hire.age, 
                       "max.yos" = max.yos, "max.age" = 80,
                       "reduced.yos" = NA, "reduced.age" = NA, 
                       "reduced.yos.age" = 80, "payroll_growth_assumption" = payroll_growth_assumption,
                       "normal.yos" = 30, "normal.age" = 67, "normal.yos.age" = 80, 
                       "normal.yos.age.min" = 65,
                       "start.wage" = start.wage*1000, "fas" = 5, "arr" = arr)
  
  
  #View(data.nm.erb)
  #View(data.nm.erb$arr)
  
  #####  Project Payroll
  payroll <- matrix(NA,manual$max.yos,5) 
  colnames(payroll) <- c("YOS", "AGE", "SALARY", "FAS", "MERIT")
  
  payroll[,1] <- seq(0, manual$max.yos-1, by = 1)
  payroll[,2] <- seq(manual$hire.age, (manual$hire.age+manual$max.yos-1), by = 1)
  
  
  payroll[1,3] <- as.numeric(manual$start.wage)*1000
  payroll[1,2] <- as.numeric(manual$hire.age)
  manual$payroll_growth_assumption <- as.numeric(manual$payroll_growth_assumption)
  payroll <- as.data.frame(payroll)
  
  ### Manual Merit payroll increase rates (can substitute w/ Excel sheet)
  payroll[1:15,5]<- c(7.00,3.50,2.75,2.25,1.75,1.50,1.25,1.00,0.75,0.50,0.25,0.25,0.25,0.25,0.25)
  payroll[16:manual$max.yos,5]<- c(0)
  
  ## Calculate Payroll growth & populate YOS & AGE columns
  
  for(i in (2:manual$max.yos)){
    payroll[i,3] <- payroll[i-1,3]*(1+(manual$payroll_growth_assumption + (payroll[i,5]/100)))
  }
  #View(payroll)
  
  #### Final Average Salary ####
  
  smooth <- manual$smooth
  payroll[(as.numeric(smooth)+1),4] <- mean(payroll[1:smooth,3])
  
  ## Calculate Final Average Salary
  
  for(i in (smooth:(manual$max.yos-2))){
    payroll[i+2,4] <- mean(payroll[(i-3):(i+1),3])
  }
  
  payroll
  
}

##### Pulling available data from the database

##### Final Average Salary (FAS)

#### components: 
#Starting Salary, 
#Wage Inflation/Payroll Growth Assumption
#Merit/Promotion Growth Rate
#Smoothing of FAS (in Years)

## Total Salary Growth Rate(t-1) = Payroll Growth %(t-1) + Merit Growth %(t-1)
## Salary(t) = Salary(t-1) * (1 + Total Salary Growth Rate(t-1))
## Final Average Salary(t) + Average(Salary(t-SmoothedYears):Salary(t-1))

#SHINY
      #ADD slider input to choose year range
      sliderInput('salary', 'Starting Salary ($ thousands)', min = 20, max = 50, value = 30, step = 2.5)
      selectInput("salary.growth", "Salary Growth Rate (%)", choices = c(2, 2.5, 3, 3.5,4), 
                  selected = 3)
      selectInput("smooth", "FAS Smoothing (Years)", choices = c(3, 4, 5), 
                  selected = 5)      


#wrap around secure_app fora password protection
##########################
######Shiny app[server] -------------------------------------------------

  
  
  ##Create a reactive datapull object to use for shiny graphics later
  

  #Create interactive plot
  plotly::renderPlotly({
    
    data <- pullStateData(2001)
    data <- filterData(data, 2017)
    data <- data %>% select(year, plan_name, state, arr,wage_inflation,payroll_growth_assumption,ee_nc_pct)
    data <- data %>% filter(plan_name == "New Mexico Educational Retirement Board")
    data.nm.erb <- as.data.table(data)
    data.nm.erb <- data.nm.erb[year == 2019] 
    
    data.nm.erb$arr <- last(na.omit(data$arr))
    data.nm.erb$wage_inflation <- last(na.omit(data$wage_inflation))
    data.nm.erb$payroll_growth_assumption <- last(na.omit(data$payroll_growth_assumption))
    data.nm.erb$ee_nc_pct <- last(na.omit(data$ee_nc_pct))
    
    
        #####  Using fas function + Adding manual assumptions
    
    graph <- data.frame(fas(arr = 0.07, 
                 payroll_growth_assumption = as.numeric(input$salary.growth)/100, start.wage = input$salary, 
                 max.yos = 50, hire.age = 25, smooth = input$smooth))
    
    graph <- data.frame(graph)
    #View(graph$YOS)
    #Call on a reactive data that was filtered above
    
    #View(UAL)
    #Graphics manual: https://bbc.github.io/rcookbook/
    #https://github.com/ReasonFoundation/pensionviewr/blob/master/README.Rmd
    
    p <- ggplot() +
      ggtitle(label = paste0("Final Average Salary"))+
      
      geom_line(data=graph, aes(x=YOS, y=FAS, 
                              color="Final Average Salary"),
                size = 1.00)+
      
    scale_colour_manual(values=c(palette_reason$Orange))+
      scale_y_continuous(labels = function(x) paste0("$",x), name = "")+
      scale_x_continuous(labels = function(x) paste0(x, ""), name = "Years of Service")+
      theme_bw()
    #Annotating ending UAL
   
    p <- ggplotly(p)
    p
  })
#rsconnect::appDependencies()
```

**Components**: </p>
- Starting Salary </p>
- Wage Inflation/Payroll Growth Assumption</p>
- Merit/Promotion Growth Rate</p>
- Smoothing of FAS (in Years)

**Calculations**: 
$$\begin{equation}
* Total Salary Growth Rate_{(t-1)} = Payroll Growth Rate_{(t-1)} + Merit Growth Rate_{(t-1)}
\end{equation}$$
$$\begin{equation}
* Salary_{(t0)} = Starting Salary ($)
\end{equation}$$
$$\begin{equation}
* Salary_{(t)} = Salary_{(t-1)} * (1 + Total Salary Growth Rate_{(t-1)})
\end{equation}$$
$$\begin{equation}
* Final Average Salary_{(t)} + \sum_{i=1}^{n}{\frac{(Salary_{(t-SmoothedYears)}:Salary_{(t-1)})}{n}}
\end{equation}$$

## Mortality Rates

```{r , echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
 
###Load/install packages
#R.Version()
#https://github.com/ReasonFoundation/pensionviewr
#Create token -> usethis::edit_r_environ() -> restart -> Sys.getenv("GITHUB_PAT")
#install.packages('devtools')
#library(devtools)
#devtools::install_github("ReasonFoundation/reasontheme",force = TRUE)
#devtools::install_github("ReasonFoundation/pensionviewr", force = TRUE)
library(reasontheme)
library(pensionviewr)
library(ggplot2)
library(tidyverse)
library(tseries)
library(data.table)
library(readr)
library(rsconnect)
library(dplyr)
library(plyr)
library(reasontheme)
library(pensionviewr)
#library(janitor)
library(tidyverse)
library(plyr)
#library(openxlsx)
library(tseries)
#library(ggplot2)
library(data.table)
library(openxlsx)
#library(readr)
library(rsconnect)
library(base64enc)
#Shiny-----------
library(shiny)
library(shinyWidgets)
library(shinymanager)
library(repmis)
#library(shinyFiles)
library(DT)
library(plotly)
library(shiny)

###### Mortality #####
manual <- data.table("vesting" = 5, "smooth" = 5, "hire.age" = 25, "max.yos" = 56, "max.age" = 80,
                     "reduced.yos" = NA, "reduced.age" = NA, "reduced.yos.age" = 80, "payroll.growth" = 0.03,
                     "normal.yos" = 30, "normal.age" = 65, "normal.yos.age" = 80, "normal.yos.age.min" = 65,
                     "start.wage" = 30000, "fas" = 5)

#Loading RP-2014 Mortality Table (SOA)
urlfile="https://raw.githubusercontent.com/ReasonFoundation/databaseR/master/files/RP-2014.csv"
rp.2014 <- data.table(
  read_csv(url(urlfile), col_names = TRUE, na = c(""),skip_empty_rows = TRUE, col_types = NULL)
)

#Loading Pub-2010 Teachers Mortality Table (SOA)
urlfile="https://raw.githubusercontent.com/ReasonFoundation/Pension-Modeling/master/data/Pub-2010.Mortality.Teachers.csv"
pub.2010 <- data.table(
  read_csv(url(urlfile), col_names = TRUE, na = c(""),skip_empty_rows = TRUE, col_types = NULL)
)

#Create extended table to add "smoothed" mortality rates between Pub-2010 given ones

#View(pub.2010)
pub.2010.smooth  <- matrix(NA,96,3) 
pub.2010.smooth[,1] <- seq(25, 120, by = 1)
pub.2010.smooth <- data.frame(pub.2010.smooth)

#Add w/ given Pub-2010 rates
pub.2010.smooth[1,2]  <- pub.2010[1,2]
pub.2010.smooth[1,3]  <- pub.2010[1,3]
for(i in (1:51)){
  pub.2010.smooth[(5*i+1),2]  <- pub.2010[(i+1),2]
  pub.2010.smooth[(5*i+1),3]  <- pub.2010[(i+1),3]
}

colnames(pub.2010.smooth) <- c("age", "male", "female")

## Compound function
#compound <- function(initial, rate, period){
#  value <- matrix(NA,period,1) 
#  for(i in (1:period)){
#  value[i,] <- initial*(1+rate)^(i)
#  }
#  value
#}

## Addition function
add <- function(initial, amount, period){
  value <- matrix(NA,period,1) 
  for(i in (1:period)){
    value[i,] <- initial+(amount)*(i)
  }
  value
}

#### Create Smoothed Rates for Males ####

#Create containers
scale <- matrix(NA,(length(pub.2010$male)-1),1) 
value <- matrix(NA,4,(length(pub.2010$male)-1)) 

#CALCULATE INCREMENTAL smoothing amounts
#And use the, to Create 4-year scale to each of Pub-2010 given mortality rates

for(k in (1:(length(pub.2010$male)-1))){
  #Calculating the proportional change within 4-year window (dividing by 5)
  scale[k,] <- (pub.2010$male[k+1]-pub.2010$male[k])/5
  #Applying this change to previous year Mortality rate for smoothing 
  value[,k] <- add(pub.2010$male[k], scale[k],4)
}

value <- data.frame(value)

#Putting each column under each other 
#for a full set of smoothed Mortality rate in 1 column

value2 <- rbind(value[,1], value[,2]) 
value2 <- rbind(value2, value[,3]) 
value2 <- rbind(value2, value[,4]) 
value2 <- rbind(value2, value[,5]) 
value2 <- rbind(value2, value[,6]) 
value2 <- rbind(value2, value[,7]) 
value2 <- rbind(value2, value[,8]) 
value2 <- rbind(value2, value[,9]) 

value2 <- data.frame(value2)
value <- pivot_longer(value2,X1:X4)
#View(value)
#View(pub.2010.smooth2)
#Finding at which ages we have missing Mortality rates
#Or ages we apply smoothing for
pub.2010.smooth2 <- pub.2010.smooth %>%
  filter(is.na(male))
pub.2010.smooth2 <- pub.2010.smooth2[1:36,1]

x <- cbind(pub.2010.smooth2, value[,2])
colnames(x) <- c("age", "male")
x <- data.frame(x)

##Attaching Smoother Rates to Missing Age column
pub.2010 <- data.frame(pub.2010)
#View(x)
#View(pub.2010.smooth[,1:2])
#View(x)
#Joining
pub.2010.smoothed_male <- pub.2010.smooth[,1:2] %>%
  ## Join Original Pub-2010 rates w/ Smoothed Rate column
  full_join(x, by = c("age"), suffix = c("_Pub_2010", "_Smoothed")) %>%
  ## Replace all NAs w/ 0
  replace_na(list("male_Pub_2010" = 0,
                  "male_Smoothed" = 0)) %>%
  ## Add rates in both columns to create one merged column
  mutate(Pub2010Smoothed = male_Pub_2010+male_Smoothed) %>%
  ## Keep age & the new column
  select(age, Pub2010Smoothed)


#### Create Smoothed Rates for Females ####

#Create containers
scale <- matrix(NA,(length(pub.2010$female)-1),1) 
value <- matrix(NA,4,(length(pub.2010$male)-1)) 

#CALCULATE INCREMENTAL smoothing amounts
#And use the, to Create 4-year scale to each of Pub-2010 given mortality rates

for(k in (1:(length(pub.2010$female)-1))){
  #Calculating the proportional change within 4-year window (dividing by 5)
  scale[k,] <- (pub.2010$female[k+1]-pub.2010$female[k])/5
  #Applying this change to previous year Mortality rate for smoothing 
  value[,k] <- add(pub.2010$female[k], scale[k],4)
}

#View(value)
value <- data.frame(value)
#Putting each column under each other 
#for a full set of smoothed Mortality rate in 1 olumn

value2 <- rbind(value[,1], value[,2]) 
value2 <- rbind(value2, value[,3]) 
value2 <- rbind(value2, value[,4]) 
value2 <- rbind(value2, value[,5]) 
value2 <- rbind(value2, value[,6]) 
value2 <- rbind(value2, value[,7]) 
value2 <- rbind(value2, value[,8]) 
value2 <- rbind(value2, value[,9]) 

value2 <- data.frame(value2)
value <- pivot_longer(value2,X1:X4)
#View(value)
#View(pub.2010.smooth2)
#Finding at which ages we have missing Mortality rates
#Or ages we apply smoothing for
pub.2010.smooth2 <- pub.2010.smooth %>%
  filter(is.na(female))
pub.2010.smooth2 <- pub.2010.smooth2[1:36,1]

x <- cbind(pub.2010.smooth2, value[,2])
colnames(x) <- c("age", "female")
x <- data.frame(x)
##Attaching Smoother Rates to Missing Age column
pub.2010 <- data.frame(pub.2010)
#View(x)
#View(pub.2010.smooth[,1:2])

#Joining
pub.2010.smooth <- pub.2010.smooth %>% select(age, female)
pub.2010.smoothed_female <-  pub.2010.smooth %>%
  ## Join Original Pub-2010 rates w/ Smoothed Rate column
  full_join(x, by = c("age"), suffix = c("_Pub_2010", "_Smoothed")) %>%
  ## Replace all NAs w/ 0
  replace_na(list("female_Pub_2010" = 0,
                  "female_Smoothed" = 0)) %>%
  ## Add rates in both columns to create one merged column
  mutate(Pub2010Smoothed = female_Pub_2010+female_Smoothed) %>%
  ## Keep age & the new column
  select(age, Pub2010Smoothed)

####################### RESULTS
#View(pub.2010.smoothed_male)
#View(pub.2010.smoothed_female)

#Adjust the Smoothed Estimated/Add smoothed rates in-between Pub-2010 rates
#View(rp.2014)
urlfile="https://raw.githubusercontent.com/ReasonFoundation/databaseR/master/files/MP_2017_Male.csv"
mp.2017.male <- read_csv(url(urlfile), col_names = TRUE, na = c(""),skip_empty_rows = TRUE, col_types = NULL)
mp.2017.male <- mp.2017.male[1:101,]

urlfile="https://raw.githubusercontent.com/ReasonFoundation/databaseR/master/files/MP_2017_Female.csv"
mp.2017.female <- read_csv(url(urlfile), col_names = TRUE, na = c(""),skip_empty_rows = TRUE, col_types = NULL)
mp.2017.female <- mp.2017.female[1:101,]

#View(mp.2017.male)

#[1] Project Pub2010 from 2010 w/ MP-2017
#[2] Project RP-2014 from 2014 w/ MP-2017
pub2010.male.scale <- mp.2017.male 
pub2010.male.scale <- pub2010.male.scale[1:51,] %>% filter(Age >= 25)
pub2010.male.scale <- pub2010.male.scale[,-(2:11)] 
pub2010.male.scale.2033 <- pub2010.male.scale$`2033`

pub.2010.smoothed_male <- pub.2010.smoothed_male[1:46,]
pub2010.male.scale$`2010` <- pub.2010.smoothed_male$Pub2010Smoothed

## Take 2010 smotthed rates & multiply by (1-MP-2017 rate for 2011...)
for (k in (3:length(pub2010.male.scale))){
  pub2010.male.scale[,k] <- pub2010.male.scale[,as.numeric(k-1)]*(1-pub2010.male.scale[,k])
  
}

#View(pub2010.male.scale)

pub2010.male.scale <- data.table(pub2010.male.scale)
x <-seq(2034, 2130, by = 1)

## Using data.table to add new columns
## Using adjusted 2033 pub-2010 rates & multiply them by (1-MP-2017 for 2033)^(distance from 2033)
for (k in (first(x):last(x))){
  pub2010.male.scale[, as.character(k) := pub2010.male.scale$`2033`*(1-pub2010.male.scale.2033)^(k-2033)]
}


pub2010.female.scale <- mp.2017.female 
pub2010.female.scale <- pub2010.female.scale[1:51,] %>% filter(Age >= 25)
pub2010.female.scale <- pub2010.female.scale[,-(2:11)] 
pub2010.female.scale.2033 <- pub2010.female.scale$`2033`

pub.2010.smoothed_female <- pub.2010.smoothed_female[1:46,]
pub2010.female.scale$`2010` <- pub.2010.smoothed_female$Pub2010Smoothed

## Take 2010 smotthed rates & multiply by (1-MP-2017 rate for 2011...)
for (k in (3:length(pub2010.female.scale))){
  pub2010.female.scale[,k] <- pub2010.female.scale[,as.numeric(k-1)]*(1-pub2010.female.scale[,k])
  
}

#View(pub2010.male.scale)

pub2010.female.scale <- data.table(pub2010.female.scale)
x <-seq(2034, 2130, by = 1)

## Using data.table to add new columns
## Using adjusted 2033 pub-2010 rates & multiply them by (1-MP-2017 for 2033)^(distance from 2033)
for (k in (first(x):last(x))){
  pub2010.female.scale[, as.character(k) := pub2010.female.scale$`2033`*(1-pub2010.female.scale.2033)^(k-2033)]
}

#View(pub2010.male.scale)
#View(pub2010.female.scale)

pub2010.male.scale[,2:122] <- round(pub2010.male.scale[,2:122]*100,5)
pub2010.female.scale[,2:122] <- round(pub2010.female.scale[,2:122]*100,5)

### Visualize probability of mortality next year (male vs. female)
## Pub-2010
pub2010.male.scale <- as.data.frame(pub2010.male.scale)

graph <- ggplot(pub2010.male.scale, aes(x = pub2010.male.scale[,1]))+
  ggtitle(label = paste0("Pub-2010 Teacher Mortailty Next Year <br> (Actives Age 25-70)"))+
    geom_line(aes(y = diag(as.matrix(pub2010.male.scale[,-1])),group = 1,
                text = paste("Mortality Next Year (Male):",                                            round(diag(as.matrix(pub2010.male.scale[,-1])),4), "%")),
              color = palette_reason$LightBlue, size = 1)+
  geom_line(aes(y = diag(as.matrix(pub2010.female.scale[,-1])), group = 1,
                text = paste("Mortality Next Year (Female):",
                       round(diag(as.matrix(pub2010.female.scale[,-1])),4), "%")),
            color = palette_reason$LightOrange, size = 1)+
  scale_y_continuous(labels = function(x) paste0(x,"%"), name = "Pub-2010 Mortality Rates")+theme_bw()+
  scale_x_continuous(labels = function(x) paste0(x), name = "Age")

graph <- ggplotly(graph, tooltip = c("text"))
graph <-graph %>% layout(autosize = TRUE, legend = list(orientation = "v", x=0.01, y = 1))
graph
```

**Components**: </p>
- Actives Mortality Table (e.g. RP-2014, plan's own table) </p>
- Retired Mortality Table (set Normal eligibility age) </p>
- Any Adjustments Applied (e.g. set back 1 year) </p>
- Improvement Sale Used (e.g. MP-2017) & Year Projections Start </p>


**Calculations**: 
$$\begin{equation}
* Pub2010^{MortalityAdjusted}_{(t)} = Pub2010^{MortalityRate}_{(t-1)} * (1-MP2017^{AdjustmentRate}_{(t-1)})
\end{equation}$$
$$\begin{equation}
* Pub2010^{CompoundSurvival}_{(t)} = Pub2010^{CompoundSurvival}_{(t-1)} * (1-Pub2010^{MortalityAdjusted}_{(t-1)})
\end{equation}$$

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.



